---
title: "Assignment Report"
author: "Muhammad Hafiz Hasan"
date: "January 31, 2016"
output: html_document
---
# Executive Summary
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

The aim of this report is to predict how well the six participants in doing the exercises based on data from accelerometers placed on the belt, forearm, arm and dumbell.

### Libraries
Following are the required libraries
```{r}
library(caret)
library(corrplot)
library(kernlab)
library(knitr)
library(randomForest)
```
```{r setoptions, echo = FALSE}
opts_chunk$set(cache = FALSE)
```

### Loading and preprocessing the data
Two csv files contatining the training and test data for this analysis purposes. 

```{r, eval = FALSE}
# check if a data folder exists; if not then create one
if (!file.exists("data-source")) {dir.create("data-source")}

# file URL and destination file
training_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
training_ds <- "./data-source/pml-training.csv"
testing_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testing_ds <- "./data-source/pml-testing.csv"

# download the file and note the time
download.file(training_url, destfile = training_ds)
download.file(testing_url, destfile = testing_ds)
date_downloaded <- date()
```

The training data was then loaded into R.

```{r}
# read the csv file for training 
data_training <- read.csv("./data-source/pml-training.csv", na.strings= c("NA",""," "))
```

Remove all first eight identifiers columns. Columns with alot of NA values also remove to minimize noise for this experiment.

```{r}
# clean the data by removing columns with NAs etc
data_training_NAs <- apply(data_training, 2, function(x) {sum(is.na(x))})
data_training_clean <- data_training[,which(data_training_NAs == 0)]

# remove identifier columns such as name, timestamps etc
data_training_clean <- data_training_clean[8:length(data_training_clean)]
```

### Creating a model
The test data set was then split up into 2 parts with ratio 70:30 for training and cross validation in order to train the model and then test it against data it was not specifically fitted to.

```{r}
# split the cleaned testing data into training and cross validation
in_train <- createDataPartition(y = data_training_clean$classe, p = 0.7, list = FALSE)
training <- data_training_clean[in_train, ]
crossval <- data_training_clean[-in_train, ]
```

A random forest model was selected to predict the classification because it has methods for balancing error in class population unbalanced data sets. The correlation between any two trees in the forest increases the forest error rate. Therefore, a correllation plot was produced in order to see how strong the variables relationships are with each other.

```{r, fig.height = 6, fig.width = 8}
# plot a correlation matrix
correlation_matrix <- cor(training[, -length(training)])
corrplot(correlation_matrix, order = "FPC", method = "circle", type = "lower", tl.cex = 0.8,  tl.col = rgb(0, 0, 0))
```

In this type of plot the dark red and blue colours indicate a highly negative and positive relationship respectively between the variables. There isn't much concern for highly correlated predictors which means that all of them can be included in the model.

Then a model was fitted with the outcome set to the training class and all the other variables used to predict.

```{r}
# fit a model to predict the classe using everything else as a predictor
model <- randomForest(classe ~ ., data = training)
model
```

The model produced a very small OOB error rate of .56%. This was deemed satisfactory enough to progress the testing.

### Cross-validation
The model was then used to classify the remaining 30% of data. The results were placed in a confusion matrix along with the actual classifications in order to determine the accuracy of the model.

```{r}
# crossvalidate the model using the remaining 30% of data
predict_cross_value <- predict(model, crossval)
confusionMatrix(crossval$classe, predict_cross_value)
```

This model yielded a 99.3% prediction accuracy. Again, this model proved very robust and adequete to predict new data.

### Predictions
A separate data set was then loaded into R and cleaned in the same manner as before. The model was then used to predict the classifications of the 20 results of this new data.

```{r}
# apply the same treatment to the final testing data
data_test <- read.csv("./data/pml-testing.csv", na.strings= c("NA",""," "))
data_test_NAs <- apply(data_test, 2, function(x) {sum(is.na(x))})
data_test_clean <- data_test[,which(data_test_NAs == 0)]
data_test_clean <- data_test_clean[8:length(data_test_clean)]

# predict the classes of the test set
prediction_test <- predict(model, data_test_clean)
prediction_test
```

### Conclusions
With the abundance of information given from multiple measuring instruments it's possible to accurately predict how well a person is preforming an excercise using a relatively simple model. 